---
title: "Course Project 1"
author: "Nigel Millick"
---

## Step 1

Loading and preprocessing the data 
```{r}
data <- "repdata%2Fdata%2Factivity.zip"

if(!file.exists("~/R/win-library/repdata%2Fdata%2Factivity")) {
        unzip(data)
}

activity <- read.csv("~/R/win-library/repdata%2Fdata%2Factivity/activity.csv")
```


## Step 2

What is mean total number of steps taken per day?
```{r} 
stepsByDay <- aggregate(steps ~ date, activity, sum)

hist(stepsByDay$steps, breaks = 20, main = "Total Steps Per Day", col="green", xlab="Number of Steps")

meanStep <- mean(stepsByDay$steps)
medianStep <- median(stepsByDay$steps)
```

The `mean` is `r meanStep` and the `median` is `r medianStep`.

## Step 3

What is the average daily activity pattern?
```{r}
stepsByInterval <- aggregate(steps ~ interval, activity, mean)

plot(stepsByInterval$interval,stepsByInterval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

maxInterval <- stepsByInterval[which.max(stepsByInterval$steps),1]
```

The 5-minute interval containing the maximum number of steps is `r maxInterval`.

## Step 4
Imputing missing values

```{r}
activity2 <- activity
incomplete <- is.na(activity2$steps)

averageStep <- tapply(activity2$steps, activity2$interval, mean, na.rm = TRUE, simplify = T)
activity2$steps[incomplete] <- averageStep[as.character(activity2$interval[incomplete])]
```

The NA's were replaced by the average over the 5 minute intervals for the period. This helps ensure smoothness across the graph.

Recount total steps by day and create Histogram. 
```{r}
stepsByDay2 <- aggregate(steps ~ date, activity2, sum)
hist(stepsByDay2$steps, breaks = 20, main = "Total Steps per Day", col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(stepsByDay$steps,breaks = 20, main = "Total Steps per Day", col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Calculate new mean and median for imputed data. 
```{r}
meanStep2 <- mean(stepsByDay2$steps)
medianStep2 <- median(stepsByDay2$steps)
```

Calculate difference between imputed and non-imputed data.
```{r}
meanDif <- meanStep2 - meanStep
medDif <- medianStep2 - medianStep
```

Calculate total difference.
```{r}
totalDif <- sum(stepsByDay2$steps) - sum(stepsByDay$steps)
```
* The imputed data mean is `r stepMean2`
* The imputed data median is `r stepMedian2`
* The difference between the non-imputed mean and imputed mean is `r meanDif`
* The difference between the non-imputed mean and imputed mean is `r medDif`
* The difference between total number of steps between imputed and non-imputed data is `r totalDif`.
* The means and medians appear roughly the same after imputing the missing data


## Step 5
Are there differences in activity patterns between weekdays and weekends? 
``` {r}

activity2$week = as.factor(ifelse(is.element(weekdays(as.Date(activity2$date)),weekdays), "Weekday", "Weekend"))

stepsByInterval2 <- aggregate(steps ~ interval + week, activity2, mean)

library(lattice)

xyplot(stepsByInterval2$steps ~ stepsByInterval2$interval | stepsByInterval2$week, 
       main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```

There appears to be more activity on weekends than weekdays.